---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  fig.retina = 3
)

library(emoji)

manuscript_target <- targets::tar_read(manuscript)

yaml_metadata <- rmarkdown::yaml_front_matter(here::here("manuscript", "manuscript.qmd"))
title <- yaml_metadata$title
abstract <- yaml_metadata$abstract
```

# `r title`

[Suparna Chaudhry](https://www.suparnachaudhry.com/) • Lewis and Clark College
[Audrey Comstock](https://audreylcomstock.weebly.com/) • Arizona State University
[Andrew Heiss](https://www.andrewheiss.com/) • Andrew Young School of Policy Studies • Georgia State University

<!-- badges: start -->
<!-- badges: end -->

---

## Abstract

`r abstract`

---

## `r emoji_glue(":lemon::devil:")`: Note on "lemon lucifer" project name

Because project titles change all the time with revisions, rewriting, and peer review, we used [{codename}](http://svmiller.com/codename/) to generate an internal-to-us project name that won't change.

```{r generate-code-name, eval=FALSE}
library(codename)
codename_message()
#> code name generated by {codename} v.0.5.0. R version 4.4.0 (2024-04-24).

codename(type = "gods", seed = "derogate backslides")
#> [1] "lemon lucifer"
```


## Licenses

**Text and figures:** All prose and images are licensed under Creative Commons ([CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)).

**Code:** All code is licensed under the [MIT License](LICENSE.md).

## Contributions and Code of Conduct

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
